Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-05 paper.qmd
Updated Figure 9

num_bins_x_two_curvy <- 22

@@

tsne_data_gau <- read_rds("data/five_gau_clusters/tsne_data_five_gau_71.rds")
gau1_scaled_obj <- gen_scaled_data(
  data = tsne_data_gau)
tsne_gau_scaled <- gau1_scaled_obj$scaled_nldr

tsne_gau <- tsne_gau_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2)) +
  geom_point(alpha=0.3) +
  geom_rect(aes(xmin = -0.05, xmax = 0.35, ymin = 0.28, ymax = 0.6),
               fill = "transparent", color = "grey50", linewidth = 0.8)

## Compute hexbin parameters
num_bins_x_gau1 <- 23
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

gau1_model <- fit_highd_model(
  highd_data = training_data_gau,
  nldr_data = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  q = 0.1
)

df_bin_centroids_gau1 <- gau1_model$df_bin_centroids
df_bin_gau1 <- gau1_model$df_bin

## Compute error

error_df_gau_abs <- augment(
  highd_data = training_data_gau,
  model_2d = df_bin_centroids_gau1,
  model_highd = df_bin_gau1)

error_df_gau_abs <- error_df_gau_abs |>
  mutate(sqrt_row_wise_abs_error = sqrt(row_wise_abs_error))

error_df_gau_abs <- error_df_gau_abs |>
  bind_cols(tsne_gau_scaled |>
              select(-ID))

error_df_gau_tsne <- error_df_gau_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             z = sqrt_row_wise_abs_error)) +
  stat_summary_hex(bins=15) +  
  xlim(c(0.3, 0.66)) + ylim(c(-0.01, 0.35)) +
  scale_fill_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  interior_annotation("a3", 
                      position = c(0.08, 0.9),
                      cex = 1.5)

## Triangulate bin centroids
tr1_object_gau1 <- tri_bin_centroids(
  df_bin_centroids_gau1, x = "c_x", y = "c_y")
tr_from_to_df_gau1 <- gen_edges(
  tri_object = tr1_object_gau1)

tr_df <- distinct(tibble(
  x = c(tr_from_to_df_gau1[["x_from"]], tr_from_to_df_gau1[["x_to"]]), 
  y = c(tr_from_to_df_gau1[["y_from"]], tr_from_to_df_gau1[["y_to"]])))

@@

## Hexagonal binning to have regular hexagons
hb_obj_gau1 <- hex_binning(
  data = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  q = 0.1)

tsne_data_with_hb_id <- hb_obj_gau1$data_hb_id

df_all_gau1 <- dplyr::bind_cols(training_data_gau |> dplyr::select(-ID),
                                  tsne_data_with_hb_id)

### Define type column
df <- df_all_gau1 |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset

@@

umap_data_gau <- read_rds("data/five_gau_clusters/umap_data_five_gau.rds")
gau1_scaled_obj <- gen_scaled_data(
  data = umap_data_gau)
umap_gau_scaled <- gau1_scaled_obj$scaled_nldr

umap_gau <- umap_gau_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.3, color = "#000000")

## Compute hexbin parameters
num_bins_x_gau1 <- 75
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

gau1_model <- fit_highd_model(
  highd_data = training_data_gau,
  nldr_data = umap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  q = 0.1
)

df_bin_centroids_gau1 <- gau1_model$df_bin_centroids
df_bin_gau1 <- gau1_model$df_bin

## Compute error

error_df_gau_abs <- augment(
  highd_data = training_data_gau,
  model_2d = df_bin_centroids_gau1,
  model_highd = df_bin_gau1)

error_df_gau_abs <- error_df_gau_abs |>
  mutate(sqrt_row_wise_abs_error = sqrt(row_wise_abs_error))

error_df_gau_abs <- error_df_gau_abs |>
  bind_cols(umap_gau_scaled |>
              select(-ID))

error_df_gau_umap <- error_df_gau_abs |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             z = sqrt_row_wise_abs_error)) +
  stat_summary_hex(bins=15) +  
  xlim(c(0.47, 0.5475)) + ylim(c(0.0125, 0.09)) +
  scale_fill_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  interior_annotation("b3", 
                      position = c(0.08, 0.9),
                      cex = 1.5)

## Triangulate bin centroids
tr1_object_gau1 <- tri_bin_centroids(
  df_bin_centroids_gau1, x = "c_x", y = "c_y")
tr_from_to_df_gau1 <- gen_edges(
  tri_object = tr1_object_gau1)

tr_df <- distinct(tibble(
  x = c(tr_from_to_df_gau1[["x_from"]], tr_from_to_df_gau1[["x_to"]]), 
  y = c(tr_from_to_df_gau1[["y_from"]], tr_from_to_df_gau1[["y_to"]])))

@@

## Hexagonal binning to have regular hexagons
hb_obj_gau1 <- hex_binning(
  data = umap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1)

umap_data_with_hb_id <- hb_obj_gau1$data_hb_id

df_all_gau1 <- dplyr::bind_cols(training_data_gau |> dplyr::select(-ID),
                                umap_data_with_hb_id)

### Define type column
df <- df_all_gau1 |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset

@@

pacmap_data_gau <- read_rds("data/five_gau_clusters/pacmap_data_five_gau.rds")
gau1_scaled_obj <- gen_scaled_data(
  data = pacmap_data_gau)
pacmap_gau_scaled <- gau1_scaled_obj$scaled_nldr |>
  select(PaCMAP1, PaCMAP2, ID)

pacmap_gau <- pacmap_gau_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2)) +
  geom_point(alpha=0.3, color = "#000000")

## Compute hexbin parameters
num_bins_x_gau1 <- 33
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

gau1_model <- fit_highd_model(
  highd_data = training_data_gau,
  nldr_data = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  q = 0.1
)

df_bin_centroids_gau1 <- gau1_model$df_bin_centroids
df_bin_gau1 <- gau1_model$df_bin

## Compute error

error_df_gau_abs <- augment(
  highd_data = training_data_gau,
  model_2d = df_bin_centroids_gau1,
  model_highd = df_bin_gau1)

error_df_gau_abs <- error_df_gau_abs |>
  mutate(sqrt_row_wise_abs_error = sqrt(row_wise_abs_error))

error_df_gau_abs <- error_df_gau_abs |>
  bind_cols(pacmap_gau_scaled |>
              select(-ID))

error_df_gau_pacmap <- error_df_gau_abs |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2,
             z = sqrt_row_wise_abs_error)) +
  stat_summary_hex(bins=15) + 
  xlim(c(-0.02, 0.18)) + ylim(c(0.12, 0.32)) +
  scale_fill_continuous_sequential(palette = "YlOrRd", breaks = 20) +
  interior_annotation("c3", 
                      position = c(0.08, 0.9),
                      cex = 1.5)

## Triangulate bin centroids
tr1_object_gau1 <- tri_bin_centroids(
  df_bin_centroids_gau1, x = "c_x", y = "c_y")
tr_from_to_df_gau1 <- gen_edges(
  tri_object = tr1_object_gau1)

tr_df <- distinct(tibble(
  x = c(tr_from_to_df_gau1[["x_from"]], tr_from_to_df_gau1[["x_to"]]), 
  y = c(tr_from_to_df_gau1[["y_from"]], tr_from_to_df_gau1[["y_to"]])))

@@

## Hexagonal binning to have regular hexagons
hb_obj_gau1 <- hex_binning(
  data = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  q = 0.1)

pacmap_data_with_hb_id <- hb_obj_gau1$data_hb_id

df_all_gau1 <- dplyr::bind_cols(training_data_gau |> dplyr::select(-ID),
                                pacmap_data_with_hb_id)

### Define type column
df <- df_all_gau1 |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset

@@

# write_rds(df_b, "data/five_gau_clusters/pacmap_model.rds")

>>>>

num_bins_x_two_curvy <- 24

@@

tsne_data_gau <- read_rds("data/five_gau_clusters/tsne_data_five_gau_71.rds")

## Compute hexbin parameters
num_bins_x_gau1 <- 23

algo_obj_gau1 <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_gau, 
  nldr_data = tsne_data_gau, 
  num_x_bins = num_bins_x_gau1)

tsne_gau_scaled <- algo_obj_gau1$nldr_scaled
tr_from_to_df_gau1 <- algo_obj_gau1$tr_from_to_df
df_bin_centroids_gau1 <- algo_obj_gau1$df_bin_centroids
df_bin_gau1 <- algo_obj_gau1$df_bin

@@

df_bin_centroids_gau1 <- df_bin_centroids_gau1 |>
  dplyr::filter(std_counts > 0)

@@

umap_data_gau <- read_rds("data/five_gau_clusters/umap_data_five_gau.rds")

## Compute hexbin parameters
num_bins_x_gau1 <- 75
algo_obj_gau1 <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_gau, 
  nldr_data = umap_data_gau, 
  num_x_bins = num_bins_x_gau1)

umap_gau_scaled <- algo_obj_gau1$nldr_scaled
tr_from_to_df_gau1 <- algo_obj_gau1$tr_from_to_df
df_bin_centroids_gau1 <- algo_obj_gau1$df_bin_centroids
df_bin_gau1 <- algo_obj_gau1$df_bin

@@

df_bin_centroids_gau1 <- df_bin_centroids_gau1 |>
  dplyr::filter(std_counts > 0)

@@

pacmap_data_gau <- read_rds("data/five_gau_clusters/pacmap_data_five_gau.rds")

## Compute hexbin parameters
num_bins_x_gau1 <- 33

algo_obj_gau1 <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_gau, 
  nldr_data = pacmap_data_gau, 
  num_x_bins = num_bins_x_gau1)

pacmap_gau_scaled <- algo_obj_gau1$nldr_scaled
tr_from_to_df_gau1 <- algo_obj_gau1$tr_from_to_df
df_bin_centroids_gau1 <- algo_obj_gau1$df_bin_centroids
df_bin_gau1 <- algo_obj_gau1$df_bin

@@

df_bin_centroids_gau1 <- df_bin_centroids_gau1 |>
  dplyr::filter(std_counts > 0)

@@
